Fake News vs Fox News

### 2024-04-18 paper/Fake News vs Fox News.qmd
fix abstract

---
title: "Fake News vs Fox News"
subtitle: "Exploring the Influence of Media Preferences on Voting Behavior in the 2020 U.S. Presidential Election Among Voters from Different Parties"
author: 
  - Hannah Yu
thanks: "Code and data are available at: https://github.com/hannahyu07/Fox-News"
date: today
date-format: long
abstract: "This study examines the impact of media consumption on voting behavior in the 2020 U.S. presidential election with data from the Cooperative Election Study (CES 2020). Using logistic regression, I analyze the relationship between watching specific media networks and voting for either Trump or Biden among individuals with varying political affiliations. The analysis shows that viewing political leaning networks have a significant impact on voting preferences among Democrats, Republicans, and Independents, with CNN viewers more likely to support Biden and Fox News viewers favoring Trump. The findings that among voters from all parties, Independent voters' supports are heavily determined by their preferred media emphasize media's role in shaping electoral outcomes and the importance of promoting media literacy and addressing challenges created by media polarization"
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---
```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(here)
library(rstanarm)
library(modelsummary)
library(ggplot2)
library(knitr)
library(marginaleffects)
library(plotly)
library(tibble)
library(margins)
library(testthat)
library(kableExtra)
library(arrow)
library(gridExtra)
library(dataverse)
#library(cowplot)
#library(lemon)
```

@@

The introduction is self-contained and tells a reader everything they need to know including: 1) broader context to motivate; 2) some detail about what the paper is about; 3) a clear gap that needs to be filled; 4) what was done; 5) what was found; 6) why it is important; 7) the structure of the paper. A reader should be able to read only the introduction and know what was done, why, and what was found. Likely 3 or 4 paragraphs, or 10 per cent of total.

# Introduction
The 2020 U.S. presidential election, between Republican candidate and incumbent President Donald Trump, and Democratic candidate and former Vice President Joe Biden, was one of the most intense in recent history. Trump's divisive and highly controversial statements while in office, as well as his administration's mishandling of the pandemic, widened the divide between the two parties during this election period. This division was also reflected in the media outlets favored by each party, with Fox News being known as the conservative media network highly favored by Republicans, while CNN leads as the Democrats' favorite. Often, media outlets on both sides presented differing narratives on the same events, with each side unwilling to concede.

The division of media was exacerbated by Trump's animosity against liberal networks. Trump famously labeled any unflattering issues that might work against him as "Fake News," regardless of their actual validity [@timm2020trump]. He even handed out the "Fake News Award" through Twitter to several notable liberal media outlets, including CNN [@siddiqui2018trump]. On the other hand, Democrats also have long-standing feuds with the Republican network Fox News. Biden has referred to Fox News as "one of the most destructive forces in the United States" [@stelter2022biden]. With clear political alliances and agendas, these different media outlets might subtly influence their viewers' political ideology and voting decisions. Therefore, it is important to explore whether these media networks do indeed have influences on election outcomes and, if so, the magnitude of that effect. A good understanding of this relationship could enable voters to be more critical of the media they consume.

To explore and answer these questions, this paper analyzes the influence of watching specific media on voters' voting behavior. It uses the 2020 US election data from the Cooperative Election Study (CES), which surveys Americans before and after elections. From the dataset, information on voters' media preferences on America's most prominent networks, including ABC, CNN, Fox News, etc., was gathered. Then, a Bayesian logistic regression model was employed to estimate the election outcome between Trump and Biden based on voters' media preferences and their original party affiliation. The estimand of this paper is the actual number of people who supported Trump or Biden during the 2020 election. However, due to the difficulty of collecting data on millions of voters, this paper estimates the estimand using the logistic regression model trained using a sample from the CES dataset. The findings indicate that watching media networks with specific political ideologies is associated with higher chances of voting for the candidate they prefer. Among all networks, this relationship is most apparent in CNN and Fox News, exerting heavy influence on voters who classify as "Independent" or "Other." Additionally, this paper also studies the influence of watching national newscasts on voting outcomes compared to watching both national and local newscasts, finding no significant differences between the outcomes of the two.

@@

Data for this research comes from the 2020 Cooperative Election Study (CES) [@citeCes2020], an annual US political survey. The CES contains information about Americans' political views, voting behaviours and experiences across various political geography and social contexts. 61,000 American adults participated in the survey in 2020.

@@

ces2020[1:5, 1:11] |>
  kable()
```

@@

ces2020 %>%
  select(1:6) %>%
  summary() %>%
  kable() %>%
  kable_styling(position = "left", bootstrap_options = "condensed", full_width = FALSE)

```
```{r}
#| message: false
#| echo: false
#| warning: false

ces2020 %>%
  select(7:11) %>%
  summary() %>%
  kable() %>%
  kable_styling(position = "left", bootstrap_options = "condensed", full_width = FALSE)

@@

The variable `Party` categorizes respondents into four categories. "Democrat" means the voter identifies as a Democrat; "Republican" if the voter identifies themselves as a Republican. However, in this dataset, "Independent" is not affiliated with the American Independent Party; independent here means the voter has no party affiliation. Lastly, "Other" symbolizes the voter being part of a party that is neither Democratic nor Republican. Only having four categories might overlook the nuanced and diverse political beliefs held by individuals. Some may have hybrid or unconventional political views that don't fit neatly into these categories. Therefore, their categorization might not accurately describe their party affiliations.

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix - @sec-model-details].

In this model, $y_i$ represents the binary outcome variable indicating whether an individual voted Biden (as opposed to Trump). The probability of voting for the Biden ($\pi_i$) is modeled using a logistic link function (\text{logit}($\pi_i$)), which is a linear combination of the intercept ($\alpha$) and the coefficients ($\beta_1$, $\beta_2$, $\beta_3$, $\beta_4$, $\beta_5$, $\beta_6$, $\beta_7$, $\beta_8$, $\beta_9$, $\beta_10$)) corresponding to the predictor variables media use networks, TV news type, and party affiliations, respectively. Media use networks are being represented by the following predictor variables:  \text{ABC}_i, \text{CBS}_i, \text{NBC}_i, \text{CNN}_i, \text{Fox\_News}_i, \text{MSNBC}_i, $\text{PBS}_i$, \text{Other}_i. The rest of the predictor variables are denoted as \text{TV\_type}_i and \text{Party}_i, where $i$ indexes the individuals in the dataset.

This modeling approach is chosen for several reasons. Firstly, logistic regression is well-suited for binary outcome variables, making it appropriate for analyzing voting behavior. Additionally, Bayesian methods enables the incorporation of prior knowledge and uncertainty into our analysis, which provides more robust estimates of the model parameters. Alternative modeling approaches, such as linear regression models, were also considered. However, Bayesian logistic regression is chosen to fit the resulting binary variable of voter's decision.

We run the model in R [@citeR] using the `rstanarm` package of [@citeRstanarm]. We use the default priors from `rstanarm`. Rstanarm employs Markov chain Monte Carlo (MCMC) techniques to estimate the posterior distribution of the parameters. To avoid excessive runtime, we randomly sampled 2500 observations to fit the model. Model diagnostics, including convergence checks and posterior summaries, are available in the supplementary materials (see Appendix @sec-model-details).

## Data Results
```{r}
#| echo: false
#| warning: false
#| message: false
# Function to extract legend from a ggplot object
g_legend <- function(a.gplot) {
  tmp <- ggplot_gtable(ggplot_build(a.gplot))
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  legend <- tmp$grobs[[leg]]
  return(legend)
}

@@

#Create individual plots for each news channel
plot_ABC <- ggplot(ces2020, aes(x = factor(ABC), fill = factor(voted_for))) +
  geom_bar(position = "dodge") +
  labs(title = "ABC", x = "ABC", y = "Count") +
  scale_fill_manual(values = c("red", "blue")) +  # Specify colors for Biden and Trump
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5))  # Center the title horizontally

plot_CBS <- ggplot(ces2020, aes(x = factor(CBS), fill = factor(voted_for))) +
  geom_bar(position = "dodge") +
  labs(title = "CBS", x = "CBS", y = "Count") +
  scale_fill_manual(values =  c("red", "blue")) +  # Specify colors for Biden and Trump
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5))  # Center the title horizontally

plot_NBC <- ggplot(ces2020, aes(x = factor(NBC), fill = factor(voted_for))) +
  geom_bar(position = "dodge") +
  labs(title = "NBC", x = "NBC", y = "Count") +
  scale_fill_manual(values =  c("red", "blue")) +  # Specify colors for Biden and Trump
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5))  # Center the title horizontally

plot_CNN <- ggplot(ces2020, aes(x = factor(CNN), fill = factor(voted_for))) +
  geom_bar(position = "dodge") +
  labs(title = "CNN", x = "CNN", y = "Count") +
  scale_fill_manual(values =  c("red", "blue")) +  # Specify colors for Biden and Trump
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5))  # Center the title horizontally

plot_Fox_News <- ggplot(ces2020, aes(x = factor(Fox_News), fill = factor(voted_for))) +
  geom_bar(position = "dodge") +
  labs(title = "Fox News", x = "Fox News", y = "Count") +
  scale_fill_manual(values =  c("red", "blue")) +  # Specify colors for Biden and Trump
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5))  # Center the title horizontally

plot_MSNBC <- ggplot(ces2020, aes(x = factor(MSNBC, levels = c("Yes", "No")), fill = factor(voted_for))) +
  geom_bar(position = "dodge") +
  labs(title = "MSNBC", x = "MSNBC", y = "Count") +
  scale_fill_manual(values =  c("red", "blue")) +  # Specify colors for Biden and Trump
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5))  # Center the title horizontally

plot_PBS <- ggplot(ces2020, aes(x = factor(PBS, levels = c("Yes", "No")), fill = voted_for)) +
  geom_bar(position = "dodge") +
  labs(title = "PBS", x = "PBS", y = "Count") +
  scale_fill_manual(values =  c("red", "blue")) +  # Specify colors for Biden and Trump
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5))  # Center the title horizontally

plot_Other <- ggplot(ces2020, aes(x = factor(Other, levels = c("Yes", "No")), fill = voted_for)) +
  geom_bar(position = "dodge") +
  labs(title = "Other", x = "Other", y = "Count") +
  scale_fill_manual(values = c("blue", "red")) +  # Specify colors for Biden and Trump
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5))  # Center the title horizontally

@@

@fig-network depicts the relationship between voters' voting preference and media network preference. Given that more voters supported Biden, it's evident that Biden supporters tend to watch more media networks overall. Consequently, I analyze the proportions of voters watching specific networks and their voting preferences.  Biden supporters have higher proportions of watching most networks. Among all the networks that have higher proportions of Biden voters, CNN and MSNBC, two of the most Democratic-leaning news networks, have significantly higher proportions of viewers that are Biden supporters and barely any viewers that support Trump. The only network that is being consumed by Trump supporters is Fox News, a network known to be Republican-leaning.

@@

## Model Results
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

The results, summarized in @tbl-modelresults, generally match our expectations. To avoid multicollinearity, the model excluded the category "Both" from `TV_type` and "Party Democrat" from `Party.`The intercept represents the estimated log odds of supporting Biden when all other predictors are held constant at their reference levels. In this case, the estimated log-odds of supporting Biden for individuals who do not watch any of these news networks watch both national and local newscasts and identify as a Democrat is 3.63.

@@

>>>>

---
title: "Fake News vs Fox News"
subtitle: "The Influence of Media Preferences on Voting Behavior in the 2020 U.S. Presidential Election Among Party Voters"
author: 
  - Hannah Yu
thanks: "Code and data are available at: https://github.com/hannahyu07/Fox-News"
date: today
date-format: long
abstract: "This study examines the impact of media consumption on voting behaviour in the 2020 U.S. presidential election with data from the Cooperative Election Study (CES 2020). Using logistic regression, I analyze the relationship between watching specific media networks and voting for either Trump or Biden among Democrat, Republican, Independent, and Other voters. The analysis shows that viewing political-leaning networks significantly impacts voting preferences among voters from all party backgrounds, with CNN viewers more likely to support Biden and Fox News viewers favouring Trump. The findings highlight the media’s role in shaping electoral outcomes, especially its influence on Independent voters, and the importance of addressing challenges posed by media polarization."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---
```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(here)
library(rstanarm)
library(modelsummary)
library(ggplot2)
library(knitr)
library(marginaleffects)
library(plotly)
library(tibble)
library(margins)
library(testthat)
library(kableExtra)
library(arrow)
library(gridExtra)
library(dataverse)

@@

# Introduction
The 2020 U.S. presidential election, between Republican candidate and incumbent President Donald Trump, and Democratic candidate and former Vice President Joe Biden, was one of the most intense in recent history. Trump's many highly controversial statements while in office and his administration's mishandling of the pandemic widened the divide between the two parties during this election period. This division was reflected in the media outlets favored by each party, with Fox News being known as the conservative media network highly favored by Republicans, while CNN leads as the Democrats' favorite.

The division of media was exacerbated by Trump's animosity against liberal networks. Trump famously labeled any unflattering issues that might work against him as "Fake News" [@timm2020trump]. He even handed out the "Fake News Award" through Twitter to several notable liberal media outlets, including CNN [@siddiqui2018trump]. On the other hand, Democrats also have long-standing feuds with the Republican network Fox News. Biden has referred to Fox News as "one of the most destructive forces in the United States" [@stelter2022biden]. With clear political alliances, these different media outlets might subtly influence their viewers' political ideology and voting decisions using their content. Therefore, it is important to explore whether these media networks do indeed have influences on election outcomes and, if so, the magnitude of that effect. A good understanding of this relationship could enable voters to be more critical of the media they consume.

To explore and answer these questions, this paper analyzes the influence of watching specific media on voters' voting behavior. It uses the 2020 US election data from the Cooperative Election Study (CES), which surveys Americans before and after elections @citeCes2020. From the dataset, I gathered information on voters' media preferences on America's most prominent networks, including ABC, CNN, Fox News, etc. Then, I used a Bayesian logistic regression model to estimate the election outcome between Trump and Biden based on voters' media preferences and their original party affiliation. The estimand of this paper is the actual number of people who supported Trump or Biden during the 2020 election. However, due to the difficulty of collecting data on millions of voters, this paper estimates the estimand using the logistic regression model trained using a sample from the CES dataset. The findings indicate that watching media networks with specific political ideologies is associated with higher chances of voting for the candidate they prefer. Among all networks, this relationship is most apparent in CNN and Fox News, exerting heavy influence on voters who classify as "Independent" or "Other." Additionally, this paper also studies the influence of watching national newscasts on voting outcomes compared to watching both national and local newscasts, finding no significant differences between the outcomes of the two.

@@

Data for this research comes from the 2020 Cooperative Election Study (CES) [@citeCes2020], an annual US political survey. The CES contains information about Americans' political views, voting behaviours and experiences across various political geography and social contexts. 61,000 American adults participated in the survey in 2020. This dataset is favoured over other available datasets because of its large observation numbers and wide-ranging information collected from the respondents.

@@

ces2020[1:5, 1:11] |>
  kable(booktabs = TRUE, linesep = "")
```

@@

ces2020 %>%
  select(1:6) %>%
  summary() %>%
  kable(booktabs = TRUE, linesep = "") %>%
  kable_styling(position = "left", bootstrap_options = "condensed", full_width = FALSE)

```{r}
#| message: false
#| echo: false
#| warning: false

ces2020 %>%
  select(7:11) %>%
  summary() %>%
  kable(booktabs = TRUE, linesep = "") %>%
  kable_styling(position = "left", bootstrap_options = "condensed", full_width = FALSE)

@@

The variable `Party` categorizes respondents into four categories. "Democrat" means the voter identifies as a Democrat; "Republican" if the voter identifies themselves as a Republican. However, in this dataset, "Independent" is not affiliated with the American Independent Party; independent here means the voter has no party affiliation. Lastly, "Other" symbolizes the voter being part of a party that is neither Democratic nor Republican. Only having four categories might overlook the diverse political beliefs held by individuals. Some may have hybrid or unconventional political views that don't fit neatly into these categories. Therefore, their categorization might not accurately describe their party affiliations.

## Data Analysis

```{r}
#| echo: false
#| warning: false
#| message: false
# Function to extract legend from a ggplot object
g_legend <- function(a.gplot) {
  tmp <- ggplot_gtable(ggplot_build(a.gplot))
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  legend <- tmp$grobs[[leg]]
  return(legend)
}

@@

#Create individual plots for each news channel
plot_ABC <- ggplot(ces2020, aes(x = factor(ABC), fill = factor(voted_for))) +
  geom_bar(position = "dodge") +
  labs(title = "ABC", x = "ABC", y = "Count") +
  scale_fill_manual(values = c("red", "blue")) +  # Specify colors for Biden and Trump
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(legend.position="none")

plot_CBS <- ggplot(ces2020, aes(x = factor(CBS), fill = factor(voted_for))) +
  geom_bar(position = "dodge") +
  labs(title = "CBS", x = "CBS", y = "Count") +
  scale_fill_manual(values =  c("red", "blue")) +  # Specify colors for Biden and Trump
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5))+  
  theme(legend.position="none")

plot_NBC <- ggplot(ces2020, aes(x = factor(NBC), fill = factor(voted_for))) +
  geom_bar(position = "dodge") +
  labs(title = "NBC", x = "NBC", y = "Count") +
  scale_fill_manual(values =  c("red", "blue")) +  # Specify colors for Biden and Trump
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(legend.position="none")

plot_CNN <- ggplot(ces2020, aes(x = factor(CNN), fill = factor(voted_for))) +
  geom_bar(position = "dodge") +
  labs(title = "CNN", x = "CNN", y = "Count") +
  scale_fill_manual(values =  c("red", "blue")) +  # Specify colors for Biden and Trump
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5))+
  theme(legend.position="none")

plot_Fox_News <- ggplot(ces2020, aes(x = factor(Fox_News), fill = factor(voted_for))) +
  geom_bar(position = "dodge") +
  labs(title = "Fox News", x = "Fox News", y = "Count") +
  scale_fill_manual(values =  c("red", "blue")) +  # Specify colors for Biden and Trump
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(legend.position="none")

plot_MSNBC <- ggplot(ces2020, aes(x = factor(MSNBC, levels = c("Yes", "No")), fill = factor(voted_for))) +
  geom_bar(position = "dodge") +
  labs(title = "MSNBC", x = "MSNBC", y = "Count") +
  scale_fill_manual(values =  c("red", "blue")) +  # Specify colors for Biden and Trump
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5))+
  theme(legend.position="none")

plot_PBS <- ggplot(ces2020, aes(x = factor(PBS, levels = c("Yes", "No")), fill = "Voted for")) +
  geom_bar(position = "dodge") +
  labs(title = "PBS", x = "PBS", y = "Count") +
  scale_fill_manual(values =  c("red", "blue")) +  # Specify colors for Biden and Trump
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(legend.position="none")

plot_Other <- ggplot(ces2020, aes(x = factor(Other, levels = c("Yes", "No")), fill = voted_for)) +
  geom_bar(position = "dodge") +
  labs(title = "Other", x = "Other", y = "Count") +
  scale_fill_manual(values = c("blue", "red")) +  # Specify colors for Biden and Trump
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5))+
  theme(legend.position="none")

@@

```
@fig-network depicts the relationship between voters' voting preference and media network preference. Given that more voters supported Biden, it's evident that Biden supporters tend to watch more media networks overall. Consequently, I analyze the proportions of voters watching specific networks and their voting preferences.  Biden supporters have higher proportions of watching most networks. Among all the networks that have higher proportions of Biden voters, CNN and MSNBC, two of the most Democratic-leaning news networks, have significantly higher proportions of viewers that are Biden supporters and barely any viewers that support Trump. The only network that is being consumed by Trump supporters is Fox News, a network known to be Republican-leaning.

@@

In this model, $y_i$ represents the binary outcome variable indicating whether an individual voted Biden (as opposed to Trump). The probability of voting for the Biden ($\pi_i$) is modeled using a logistic link function (\text{logit}($\pi_i$)), which is a linear combination of the intercept ($\alpha$) and the coefficients ($\beta_1$, $\beta_2$, $\beta_3$, $\beta_4$, $\beta_5$, $\beta_6$, $\beta_7$, $\beta_8$, $\beta_9$, $\beta_10$)) corresponding to the predictor variables media use networks, TV news type, and party affiliations, respectively. Media use networks are being represented by the following predictor variables:  $\text{ABC}_i$, $\text{CBS}_i$, $\text{NBC}_i$, $\text{CNN}_i$, $\text{Fox\_News}_i$, $\text{MSNBC}_i$, $\text{PBS}_i$, $\text{Other}_i$. The rest of the predictor variables are denoted as $\text{TV\_type}_i$ and $\text{Party}_i$, where $i$ indexes the individuals in the dataset.

This modeling approach is chosen for several reasons. Firstly, logistic regression is well-suited for binary outcome variables, making it appropriate for analyzing voting behavior. Additionally, Bayesian methods enables the incorporation of prior knowledge and uncertainty into the analysis, which provides more robust estimates of the model parameters. Alternative modeling approaches, such as linear regression models, were also considered. However, Bayesian logistic regression is chosen to fit the resulting binary variable of voter's decision.

I run the model in R [@citeR] using the `rstanarm` package of [@citeRstanarm]. I use the default priors from `rstanarm`. Rstanarm employs Markov chain Monte Carlo (MCMC) techniques to estimate the posterior distribution of the parameters. I randomly sampled 2500 observations to fit the model to avoid excessive run time.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

The results, summarized in @tbl-modelresults, generally match the expectations. To avoid multicollinearity, the model excluded the category "Both" from `TV_type` and "Party Democrat" from `Party.`The intercept represents the estimated log odds of supporting Biden when all other predictors are held constant at their reference levels. In this case, the estimated log-odds of supporting Biden for individuals who do not watch any of these news networks watch both national and local newscasts and identify as a Democrat is 3.63.

@@

Model diagnostics, including convergence checks and posterior summaries, and explanations can be found in Appendix @sec-model-details.
